The illusory promise of the Aligned Rank Transform

### 2025-07-07 text/40-interpretingEffects.qmd
updated appendix

**Dealing with interpretation issues.** To avoid such interpretation issues, we focus -- unless otherwise stated -- on effects that apply monotonic transformations to population distributions. This also ensures a monotonic relationship between different measures of central tendency such as medians and means (with the exception of the Cauchy distribution, where the mean is undefined). In other words, if a treatment increases the population mean, it will also increase the population median. We present an example in @fig-distributions. The figure shows two population distributions corresponding to the two intermediate levels of difficulty of our illustrative example (see @fig-example). We observe that the increased difficulty of the task translates both the population mean and the median to the right. In this case, we expect a statistical test to reject the null hypothesis, no matter whether it tests the population mean, the median, or the overall distribution shape.

@@

In conclusion, focusing solely on the issues with the rank tranformation illustrated in @fig-interactions-rank is akin to missing the forest for the trees. When parallel main effects are present, interpreting interactions can be challenging for all methods -- and ART offers no solution to this problem.

@@

GLMs are defined by a link function $g$ that connects a variable -- linearly defined by the predictors -- to a response variable $Y$, which follows an arbitrary distribution. For an experimental design with two factors, $x_1$ and $x_2$, the expected value (or mean) of $Y$ conditional on $x_1$ and $x_2$ is expressed as follows:

@@

**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects -- that is, when the coefficients of all interacting factors are non-zero. Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero. In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.

>>>>

**Dealing with interpretation issues.** To avoid such interpretation issues, we focus --- unless otherwise stated --- on effects that apply monotonic transformations to population distributions. This also ensures a monotonic relationship between different measures of central tendency such as medians and means (with the exception of the Cauchy distribution, where the mean is undefined). In other words, if a treatment increases the population mean, it will also increase the population median. We present an example in @fig-distributions. The figure shows two population distributions corresponding to the two intermediate levels of difficulty of our illustrative example (see @fig-example). We observe that the increased difficulty of the task translates both the population mean and the median to the right. In this case, we expect a statistical test to reject the null hypothesis, no matter whether it tests the population mean, the median, or the overall distribution shape.

@@

In conclusion, focusing solely on the issues with the rank tranformation illustrated in @fig-interactions-rank is akin to missing the forest for the trees. When parallel main effects are present, interpreting interactions can be challenging for all methods --- and ART offers no solution to this problem.

@@

GLMs are defined by a link function $g$ that connects a variable --- linearly defined by the predictors --- to a response variable $Y$, which follows an arbitrary distribution. For an experimental design with two factors, $x_1$ and $x_2$, the expected value (or mean) of $Y$ conditional on $x_1$ and $x_2$ is expressed as follows:

@@

**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects --- that is, when the coefficients of all interacting factors are non-zero. Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero. In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.

### 2025-07-07 text/66-Results-Interactions.qmd
updated appendix

### Interactions under parallel main effects {#interactions}
Let us now examine interaction effects when both coefficients $a_1$ and $a_2$ increase in parallel.
In this scenario, the interpretation of an interaction effect can become ambiguous. Here, we define the null hypothesis in terms of the coefficients of the latent linear model, following the recommendation of @Greene:2010 -- that is, $a_{12} = 0$.

@@

3. The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations -- regardless of the method used. Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions. As previously noted, @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model. Our findings clearly show that this is not the case.

>>>>

### Interactions under parallel main effects {#interactions}
Let us now examine interaction effects when both coefficients $a_1$ and $a_2$ increase in parallel.
In this scenario, the interpretation of an interaction effect can become ambiguous. Here, we define the null hypothesis in terms of the coefficients of the latent linear model, following the recommendation of @Greene:2010 --- that is, $a_{12} = 0$.

@@

3. The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations --- regardless of the method used. Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions. As previously noted, @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model. Our findings clearly show that this is not the case.
